Temperature and moisture effects on greenhouse gas emissions from deep active-layer boreal soils

### 2016-11-18 manuscript/response_to_reviewer_comments.tex

\medskip
{\it 1.10. Lines 70-72: These are really important considerations, and it seems appropriate to discuss them more explicitly. How is the temperature and precipitation regime of the boreal forest of interior AK expected to change? There are also indirect effects of vegetation type on soil temperatures that could be discussed here. }

We agree, although this largely echoes comments 1.7 and 1.9; please see our responses above.

\medskip
{\it 1.11. Lines 72-74: While these are important questions, they are not really addressed in this study, and so either it might make sense to leave it out, or to discuss the particulars as they apply to this study, ie: the importance and questions related to C:N ratios. }

Referee 3 made this point as well, and this sentence has been removed.

@@

We will add a sentence of caveats, noting in particular the useful but incremental nature of this study. {\bf See new lines LINES.}

% ======================================================================================
\newpage
\subsection*{Response to Referee \#3}

{\it 3.2. My main criticism is that I think that the authors over-emphasize the results of the daily emissions and that the authors should further explore (or report) the results of the controls of the cumulative C emissions. I'm curious as to whether the relationships with soil C/N and \%N observed in daily emissions still hold on cumulative emissions. The comparison between these soil parameters (i.e. ones that probably don't change much throughout the course of the incubation, including temperature) and the cumulative fluxes is perhaps more appropriate. Perhaps modelers find the controls on daily fluxes interesting and these are likely quite useful in regards to the relationship between moisture and C production (i.e. changes on a daily basis), but I think that the controls on cumulative fluxes are quite interesting and could be further explored. }

We agree that rebalancing the manuscript, focusing a bit more on controls on cumulative emissions and a bit less on the instantaneous fluxes, would strengthen it. Accordingly, we now more fully explore controls on the cumulative emissions. {\bf See new lines LINES.}

@@

\medskip
{\it 3.15. 253-254: So what variables were significant in predicting cumulative C emissions? }

Please see our response to comment 3.2 above.

\medskip
{\it 3.16. 262: First mention of vegetation stress, remove, not clear how it's related. }

We now better integrate this point, mentioning it in the introduction and clarifying its relationship to the study goals. {\bf See new lines LINES.}

@@

\medskip
{\it 3.21. 344-345: Specify that the results in Treat et al. (2015) were for anaerobic incubations and were thus likely to be much smaller. }

Thanks; we have done so. {\bf See new lines LINES.}

\medskip
{\it 3.22. 347-348: See also Lee et al. (2012)
364-365: See also Schadel et al. (2014). Also, I thought this section was a bit vague, probably could be shortened slightly. }

Thanks for the Lee et al. reference, which we had not considered (see our response to 3.7 above) but is now cited. We have also reworked and tightened section 4.2. {\bf See new lines LINES.}

### 2016-11-18 manuscript/cpcrw_manuscript.Rmd
Addressed a variety of minor comments from Reviewers 2 and 3

Rapid climatic changes, rising air temperatures, and increased fires are expected to drive permafrost degradation and alter soil carbon (C) cycling in many high-latitude ecosystems. How these soils will respond to changes in their temperature, moisture, and overlying vegetation is highly uncertain, but critical to understand given the large soil C stocks in these regions. We used a laboratory experiment to examine how temperature and moisture control CO~2~ and CH~4~ emissions from mineral soils sampled from the bottom of the annual active layer, i.e. directly above permafrost, in an Alaskan boreal forest. Gas emissions from thirty cores, subjected to two temperatures and either field moisture conditions or experimental drought, were tracked over a 100-day incubation; we also measured a variety of physical and chemical characteristics of the cores. Gravimetric water content was `r pn(gwcbegin$mean, 2)` ± `r pn(gwcbegin$sd, 2)` (unitless) at the beginning of the incubation; cores at field moisture were unchanged at the end, but drought cores had declined to `r pn(gwcendmean["Drought"], 2)` ±  `r pn(gwcendsd["Drought"], 2)`. Daily CO~2~ fluxes were positively correlated with incubation chamber temperature, core water content, and percent soil nitrogen, and had a temperature sensitivity (Q~10~) of `r pn(Q10_cumulative["Field moisture", "CO2"], 1)` and `r pn(Q10_cumulative["Drought", "CO2"], 1)` for the field moisture and drought treatments, respectively. Daily CH~4~ emissions were most strongly correlated with percent nitrogen, but neither temperature nor water content was a significant first-order predictor of CH~4~ fluxes. The cumulative production of C from CO~2~ was over `r minmag` orders of magnitudes higher than that from CH~4~. These results suggest that deep active-layer soils may be quite sensitive to changes in soil moisture, a critical factor as discontinuous permafrost thaws [COM2.5] in interior Alaska. Deep but unfrozen high-latitude soils have been shown to be strongly affected by long-term experimental warming, and these results provide insight into their future dynamics and feedback potential with future climate change.

@@

These feedbacks between vegetation type, ground cover, permafrost, and SOC may be disrupted by increased fire disturbance, particularly after more-intense fire events {Johnstone, 2010 #2465}{Genet, 2013 #4412}. Even absent disturbance, the stability of SOC is highly uncertain, as it depends on soil temperature and moisture, the ages of and ratio between the carbon (C) and nitrogen (N) pools {Weiss, 2015 #4317}{Karhu, 2014 #4124}, and its protection from competent microorganisms, enzymes, and resources {Bailey, 2012 #3624}, whether by organomineral sorption, chemical lability, or physical location {Schmidt, 2011 #3135}.

@@

Most previous studies have focused on surface soils or permafrost soils, neglecting deep active-layer soils that were identified as subject to strong effects from a two-decade warming experiment in the Alaskan Arctic {Sistla, 2013 #3666}. Such deeper soils have particular characteristics distinguishing them from both shallow active layer soils and underlying permafrost: they are most affected by interannual variability in thaw depth, potentially flipping the C source/sink status of entire ecosystems {Goulden, 1998 #50}{Harden, 2012 #4476}, and are subject to distinctive freeze-thaw, cryoturbation, and microbial dynamics, which are likely to change their sensitivity to climate change and feedback potential {Schuur, 2008 #4491}; and deep seasonally frozen soils pose particular problems for accurate modeling of high-latitude carbon dynamics {Nicolsky, 2007 #4474}. Deep soil carbon is likely to be a highly important contributor to future climate feedbacks, with modeling studies suggesting that one-third of 21^st^ century climate-induced carbon loss may originate from seasonally frozen soil carbon {Koven, 2015 #4473}. [COM1.8,2.6]

@@

These means conceal considerable variability over the course of the incubation (**Figures 2 and 3**). In the linear mixed-effects model (AIC = `r pn(AIC(step_co2_lme), 1)`), CO~2~ was positively correlated with incubation chamber temperature, core gravimetric water content, and percent soil N (all P < 0.05, and the latter two P < 0.001; **Table 2**). Temperature sensitivity decreased significantly (`r pclean(co2_temp_time)`) over the course of the incubation, while moisture sensitivity was unaffected by time. Percent C and percent N were highly correlated (r = `r pn(cor(cndata$C_percent, cndata$N_percent), 2)`) for these cores. Because percent N was a slightly stronger predictor, it was retained in the model while percent C was excluded; cf. Colman and Schimel {, 2014 #4172}. The interaction between water content and percent N was also highly significant (`r pclean(summary(step_co2_lme)$tTable["WC_gravimetric:N_percent", "p-value"])`), with high-N cores having little relationship between water content and CO~2~ flux (data not shown).

@@

Temperature and moisture senstivity was generally invariant over time: 
The cumulative production of C from CO~2~ (**Figure 4**) was over `r minmag` order of magnitudes higher than that from CH~4~, with CO~2~:CH~4~ C ratios ranging from `r minratio` million in the `r mintemp` °C "`r maxtrt`" treatment, to `r maxratio` million in the `r maxtemp` °C "`r maxtrt`" treatment. Cumulative CO~2~ evolved was highly affected by temperature (`r pclean(co2_hsd$Temperature["p adj"])`), and "field moisture" cores emitted significantly more CO~2~ than the other two moisture treatments at both temperatures (P < 0.001 for both, with no significant interactive effect). There was no difference between fluxes from the 20 °C "drought" and "controlled drought" treatments (`r pclean(co2_hsd$Treatment["Drought-Controlled drought", "p adj"])`). "Drought" cores' cumulative production was `r pn(drought_effect["4", "Reduction"], 0)`% (4 °C) and `r pn(drought_effect["20", "Reduction"], 0)`% (20 °C) lower than the cores kept at field moisture. Neither temperature (`r pclean(ch4_hsd$Temperature[,"p adj"])`) nor moisture treatment (mean `r pclean(mean(ch4_hsd$Treatment[,"p adj"]))`) was a significant factor in predicting cumulative CH~4~ fluxes.

The cumulative flux numbers above result in CO~2~ temperature sensitivity (Q~10~) values of `r pn(Q10_cumulative["Field moisture", "CO2"], 1)` and `r pn(Q10_cumulative["Drought", "CO2"], 1)` for the field moisture and drought treatments, respectively; the corresponding Q~10~ values  based on cumulative CH~4~ were `r pn(Q10_cumulative["Field moisture", "CH4"], 1)` and `r pn(Q10_cumulative["Drought", "CH4"], 1)`. Computing Q~10~ values based on fluxes normalized by water filled pore space changed these values only slightly: to `r pn(wfpsQ10_cumulative["Field moisture", "CO2"], 1)` and `r pn(wfpsQ10_cumulative["Drought", "CO2"], 1)` for CO~2~, respectively, and `r pn(wfpsQ10_cumulative["Field moisture", "CH4"], 1)` and `r pn(wfpsQ10_cumulative["Drought", "CH4"], 1)` for CH~4~.

@@

Rises in boreal air temperatures, and unpredictable precipitation changes, will warm and dry many soils, increase vegetation stress {Ju, 2016 #4320}{Barber, 2000 #1320}, degrade permafrost and deepen the active layer {Schuur, 2015 #4319}, with uncertain consequences for soil dynamics and GHG fluxes. In this laboratory experiment we found that CO~2~, but not CH~4~, fluxes from these oxic active-layer mineral soils were sensitive to temperature and, in particular, moisture.

Several studies have measured microbial respiration and GHG fluxes from soils very close to our study site. Morishita et al. {, 2014 #4316} quantified gas fluxes in the field at CPCRW and nearby forests, and found CO~2~ production to be correlated with both temperature and moisture in upland Cryosols [COM3.18], consistent with our results.[COM1.17,3.17] They found however that CH~4~ uptake (no emissions were observed) was driven by temperature only. Waldrop et al. {, 2010 #4380} incubated active-layer and permafrost soils from *Picea mariana* sites near Fairbanks, AK, under varied temperature and aerobic conditions, observing Q~10~ values of 9.0 (active layer) and 2.3 (permafrost) from -5 to 5 °C; these values are higher than we observed, consistent with the lower temperature range {Hamdi, 2013 #3664} and fundamental biokinetics {Davidson, 2006 #1669}. Waldrop et al. {, 2010 #4380} also observed flux rates of 0.001-0.10 µmol CH~4~ day^-1^ g^-1^ (~`r pn(0.001 * 16 / 0.012 / 1000, 3)`-`r pn(0.1 * 16 / 0.012 / 1000, 3)` ng C g C^-1^ day^-1^), differing by orders of magnitude between sites (but roughly similar to our observed CH~4~ emissions), and ~1-5 µg C-CO~2~ hr^-1^ g^-1^ (~`r pn(1 * 24 / 0.012, 0)`-`r pn(5 * 24 / 0.012, 0)` µg C g C^-1^ day^-1^), considerably higher than the CO~2~ rates observed from our cores. In an incubation of active-layer Alaskan permafrost peats, Treat et al. {, 2014 #4213} found CO~2~ and CH~4~ emissions to be strongly correlated with temperature and moisture. Finally, during the first 100 days of a year-long incubation of Fairbanks-area 0-10 cm mineral soils, Neff and Hooper {, 2002 #4390} observed fluxes of ~`r pn(0.2 * 1000 * 12/44, 0)`-`r pn(1.5 * 1000 * 12/44, 0)` µg C-CO~2~ g C^-1^ day^-1^, in line with the results here.

@@

discuss - changes over time (see new reuslts)
discuss - upland mixed-canopy boreal forest here; careful about extrapolation

@@

We observed very low but positive CH~4~ production from these upland mineral soils. This is contrast to many field studies that have observed CH~4~ uptake (oxidation) in dry boreal sites {Matson, 2009 #2259}{Schaufler, 2010 #4391}. Anoxic microsites in soil can however provide enough CH~4~ production to balance low-level consumption in otherwise aerobic soils {Kammann, 2009 #4424}. In addition, our results are broadly consistent with data from 65 studies summarized by Olefeldt et al. {, 2013 #3456}, who found that CH~4~ emissions were more sensitive to soil temperature in wetter ecosystems; it would have been a surprise if the little methanogenic activity in our upland, well-drained soils was temperature-sensitive at all. Methane was also a far smaller C flux than CO~2~ from these soils, in particular at higher temperatures (as CO~2~ was responsive to temperature, but CH~4~ was not). This is true more generally: for example, Treat et al. found a median CO~2~:CH~4~ production ratio of 387 for anaerobic incubations of boreal soils. This is naturally far lower than our observed aerobic (and thus high-CO~2~) ratios, but nonetheless consistent with them. Thus we see little opportunity for CH~4~ to be a significant contributor to these upland soils' C fluxes and climate feedback risk, even accounting for the stronger radiative forcing of this gas.

>>>>

Rapid climatic changes, rising air temperatures, and increased fires are expected to drive permafrost degradation and alter soil carbon (C) cycling in many high-latitude ecosystems. How these soils will respond to changes in their temperature, moisture, and overlying vegetation is highly uncertain, but critical to understand given the large soil C stocks in these regions. We used a laboratory experiment to examine how temperature and moisture control CO~2~ and CH~4~ emissions from mineral soils sampled from the bottom of the annual active layer, i.e. directly above permafrost, in an Alaskan boreal forest. Gas emissions from thirty cores, subjected to two temperatures and either field moisture conditions or experimental drought, were tracked over a 100-day incubation; we also measured a variety of physical and chemical characteristics of the cores. Gravimetric water content was `r pn(gwcbegin$mean, 2)` ± `r pn(gwcbegin$sd, 2)` (unitless) at the beginning of the incubation; cores at field moisture were unchanged at the end, but drought cores had declined to `r pn(gwcendmean["Drought"], 2)` ±  `r pn(gwcendsd["Drought"], 2)`. Daily CO~2~ fluxes were positively correlated with incubation chamber temperature, core water content, and percent soil nitrogen, and had a temperature sensitivity (Q~10~) of `r pn(Q10_cumulative["Field moisture", "CO2"], 1)` and `r pn(Q10_cumulative["Drought", "CO2"], 1)` for the field moisture and drought treatments, respectively. Daily CH~4~ emissions were most strongly correlated with percent nitrogen, but neither temperature nor water content was a significant first-order predictor of CH~4~ fluxes. The cumulative production of C from CO~2~ was over `r minmag` orders of magnitudes higher than that from CH~4~. These results suggest that deep active-layer soils may be quite sensitive to changes in soil moisture, a critical factor as discontinuous permafrost thaws in interior Alaska. Deep but unfrozen high-latitude soils have been shown to be strongly affected by long-term experimental warming, and these results provide insight into their future dynamics and feedback potential with future climate change.

@@

A number of factors may however disrupt these these feedbacks between vegetation type, ground cover, permafrost, and SOC stability. Climate changes, in particular regional warming and drying, cause vegetation stress {Ju, 2016 #4320}{Barber, 2000 #1320} and increased mortality. Climate also drives fire regime changes, and ecosystem disruption is particularly likely after more-intense fire events {Johnstone, 2010 #2465}{Genet, 2013 #4412}. Even absent disturbance, the stability of SOC is highly uncertain, as it depends on soil temperature and moisture, the ages of and ratio between the carbon (C) and nitrogen (N) pools {Weiss, 2015 #4317}{Karhu, 2014 #4124}, and its protection from competent microorganisms, enzymes, and resources {Bailey, 2012 #3624}, whether by organomineral sorption, chemical lability, or physical location {Schmidt, 2011 #3135}.

@@

Most previous studies have focused on surface soils or permafrost soils, neglecting deep active-layer soils that were identified as subject to strong effects from a two-decade warming experiment in the Alaskan Arctic {Sistla, 2013 #3666}. Such deeper soils have particular characteristics distinguishing them from both shallow active layer soils and underlying permafrost: they are most affected by interannual variability in thaw depth, potentially flipping the C source/sink status of entire ecosystems {Goulden, 1998 #50}{Harden, 2012 #4476}, and are subject to distinctive freeze-thaw, cryoturbation, and microbial dynamics, which are likely to change their sensitivity to climate change and feedback potential {Schuur, 2008 #4491}; and deep seasonally frozen soils pose particular problems for accurate modeling of high-latitude carbon dynamics {Nicolsky, 2007 #4474}. Deep soil carbon is likely to be a highly important contributor to future climate feedbacks, with modeling studies suggesting that one-third of 21^st^ century climate-induced carbon loss may originate from seasonally frozen soil carbon {Koven, 2015 #4473}.

@@

These means conceal considerable variability over the course of the incubation (**Figures 2 and 3**). In the linear mixed-effects model (AIC = `r pn(AIC(step_co2_lme), 1)`), CO~2~ was positively correlated with incubation chamber temperature, core gravimetric water content, and percent soil N (all P < 0.05, and the latter two P < 0.001; **Table 2**). Temperature sensitivity decreased significantly (`r pclean(co2_temp_time)`) over the course of the incubation, while moisture sensitivity was unaffected by time. Percent C and percent N were highly correlated (r = `r pn(cor(cndata$C_percent, cndata$N_percent), 2)`) for these cores. Because percent N was a slightly stronger predictor, it was retained in the model while percent C was excluded; cf. Colman and Schimel {, 2014 #4172}. The interaction between water content and percent N was also highly significant (`r pclean(summary(step_co2_lme)$tTable["WC_gravimetric:N_percent", "p-value"])`), although cores with N > 0.2\% exhibited little relationship between water content and CO~2~ flux (data not shown).

@@

The cumulative production of C from CO~2~ (**Figure 4**) was over `r minmag` order of magnitudes higher than that from CH~4~, with CO~2~:CH~4~ C ratios ranging from `r minratio` million in the `r mintemp` °C "`r maxtrt`" treatment, to `r maxratio` million in the `r maxtemp` °C "`r maxtrt`" treatment. Cumulative CO~2~ evolved was highly affected by temperature (`r pclean(co2_hsd$Temperature["p adj"])`), and "field moisture" cores emitted significantly more CO~2~ than the other two moisture treatments at both temperatures (P < 0.001 for both, with no significant interactive effect). There was no difference between fluxes from the 20 °C "drought" and "controlled drought" treatments (`r pclean(co2_hsd$Treatment["Drought-Controlled drought", "p adj"])`). "Drought" cores' cumulative production was `r pn(drought_effect["4", "Reduction"], 0)`% (4 °C) and `r pn(drought_effect["20", "Reduction"], 0)`% (20 °C) lower than the cores kept at field moisture. Neither temperature (`r pclean(ch4_hsd$Temperature[,"p adj"])`) nor moisture treatment (mean `r pclean(mean(ch4_hsd$Treatment[,"p adj"]))`) was a significant factor in predicting cumulative CH~4~ fluxes.

The cumulative flux numbers above result in CO~2~ temperature sensitivity (Q~10~) values of `r pn(Q10_cumulative["Field moisture", "CO2"], 1)` and `r pn(Q10_cumulative["Drought", "CO2"], 1)` for the field moisture and drought treatments, respectively; the corresponding Q~10~ values  based on cumulative CH~4~ were `r pn(Q10_cumulative["Field moisture", "CH4"], 1)` and `r pn(Q10_cumulative["Drought", "CH4"], 1)`. Computing Q~10~ values based on fluxes normalized by water filled pore space changed these values only slightly: to `r pn(wfpsQ10_cumulative["Field moisture", "CO2"], 1)` and `r pn(wfpsQ10_cumulative["Drought", "CO2"], 1)` for CO~2~, respectively, and `r pn(wfpsQ10_cumulative["Field moisture", "CH4"], 1)` and `r pn(wfpsQ10_cumulative["Drought", "CH4"], 1)` for CH~4~. Temperature and moisture senstivities were generally invariant over time: TODO.

@@

Rises in boreal air temperatures, and unpredictable precipitation changes, are changing fire disturbance regimes, warming and dry many soils, increasing vegetation stress, degrade permafrost and deepen the active layer {Schuur, 2015 #4319}, all with uncertain consequences for soil dynamics and GHG fluxes. In this laboratory experiment we found that CO~2~, but not CH~4~, fluxes from these oxic active-layer mineral soils were sensitive to temperature and, in particular, moisture.

Several studies have measured microbial respiration and GHG fluxes from soils very close to our study site. Morishita et al. {, 2014 #4316} quantified gas fluxes in the field at CPCRW and nearby forests, and found CO~2~ production to be correlated with both temperature and moisture in upland Cryosols, consistent with our results. They found however that CH~4~ uptake (no emissions were observed) was driven by temperature only. Waldrop et al. {, 2010 #4380} incubated active-layer and permafrost soils from *Picea mariana* sites near Fairbanks, AK, under varied temperature and aerobic conditions, observing Q~10~ values of 9.0 (active layer) and 2.3 (permafrost) from -5 to 5 °C; these values are higher than we observed, consistent with the lower temperature range {Hamdi, 2013 #3664} and fundamental biokinetics {Davidson, 2006 #1669}. Waldrop et al. {, 2010 #4380} also observed flux rates of 0.001-0.10 µmol CH~4~ day^-1^ g^-1^ (~`r pn(0.001 * 16 / 0.012 / 1000, 3)`-`r pn(0.1 * 16 / 0.012 / 1000, 3)` ng C g C^-1^ day^-1^), differing by orders of magnitude between sites (but roughly similar to our observed CH~4~ emissions), and ~1-5 µg C-CO~2~ hr^-1^ g^-1^ (~`r pn(1 * 24 / 0.012, 0)`-`r pn(5 * 24 / 0.012, 0)` µg C g C^-1^ day^-1^), considerably higher than the CO~2~ rates observed from our cores. In an incubation of active-layer Alaskan permafrost peats, Treat et al. {, 2014 #4213} found CO~2~ and CH~4~ emissions to be strongly correlated with temperature and moisture. Finally, during the first 100 days of a year-long incubation of Fairbanks-area 0-10 cm mineral soils, Neff and Hooper {, 2002 #4390} observed fluxes of ~`r pn(0.2 * 1000 * 12/44, 0)`-`r pn(1.5 * 1000 * 12/44, 0)` µg C-CO~2~ g C^-1^ day^-1^, in line with the results here.

@@

discuss - changes over time (see new reuslts)
discuss - upland mixed-canopy boreal forest here; careful about extrapolation
discuss - controls on cumulative emissions

@@

We observed very low but positive CH~4~ production from these upland mineral soils. This is contrast to many field studies that have observed CH~4~ uptake (oxidation) in dry boreal sites {Matson, 2009 #2259}{Schaufler, 2010 #4391}. Anoxic microsites in soil can however provide enough CH~4~ production to balance low-level consumption in otherwise aerobic soils {Kammann, 2009 #4424}. In addition, our results are broadly consistent with data from 65 studies summarized by Olefeldt et al. {, 2013 #3456}, who found that CH~4~ emissions were more sensitive to soil temperature in wetter ecosystems; it would have been a surprise if the little methanogenic activity in our upland, well-drained soils was temperature-sensitive at all. Methane was also a far smaller C flux than CO~2~ from these soils, in particular at higher temperatures (as CO~2~ was responsive to temperature, but CH~4~ was not). This is true more generally: for example, Treat et al. found a median CO~2~:CH~4~ production ratio of 387 for anaerobic incubations of boreal soils. This is naturally far lower than our observed aerobic (and thus high-CO~2~) ratios, but nonetheless consistent with them. Thus we see little opportunity for CH~4~ to be a significant contributor to these upland soils' C fluxes and climate feedback risk, even accounting for the 25x stronger radiative forcing of this gas over a 100-year time horizon {Lee, 2012 #4503}.
